---
title: "Data Preprocessing"
output: md_document
---

## Import Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

##Import CSV file
```{r cars}
nfl_df = read.csv('NFL_Spread_Data.csv')

nfl_df
```
## Create New Dataframe Containing only games afte 1999
```{r pressure, echo=FALSE}
spread_df = filter(nfl_df, schedule_season > 1999)

spread_df
```

## Replace team names that have switched cities or names since 2000
```{r pressure, echo=FALSE}
spread_df$team_home[spread_df$team_home == "Oakland Raiders"] <- "Las Vegas Raiders"
spread_df$team_home[spread_df$team_home == "Washington Redskins"] <- "Washington Commanders"
spread_df$team_home[spread_df$team_home == "Washington Football Team"] <- "Washington Commanders"
spread_df$team_home[spread_df$team_home == "St. Louis Rams"] <- "Los Angeles Rams"
spread_df$team_home[spread_df$team_home == "San Diego Chargers"] <- "Los Angeles Chargers"

spread_df$team_away[spread_df$team_away == "Oakland Raiders"] <- "Las Vegas Raiders"
spread_df$team_away[spread_df$team_away == "Washington Redskins"] <- "Washington Commanders"
spread_df$team_away[spread_df$team_away == "Washington Football Team"] <- "Washington Commanders"
spread_df$team_away[spread_df$team_away == "St. Louis Rams"] <- "Los Angeles Rams"
spread_df$team_away[spread_df$team_away == "San Diego Chargers"] <- "Los Angeles Chargers"
```

## Replace team favorite ids with team names
```{r pressure, echo=FALSE}
team_names = unique(spread_df$team_home)
team_names = append(team_names, "None")

team_ids = c("ATL", "BUF", "CLE", "DAL", "GB", "KC", "MIA", "MIN", "NE", "NO", "NYG", "LV", "PIT", "WAS", "LAR", "ARI", "BAL", "CIN", "DEN", "DET", "IND", "PHI", "LAC", "SF", "SEA", "TB", "TEN", "NYJ", "CAR", "CHI", "JAX", "HOU", "PICK")

names(team_names) = team_ids

spread_df$team_favorite_id = team_names[spread_df$team_favorite_id]

spread_df_whole = spread_df
```


## Remove unnecessary columns
```{r pressure, echo=FALSE}
spread_df = select(spread_df, c(5:10))

spread_df
```

## Create difference columns that subtracts score home and score away based on the favorite
```{r pressure, echo=FALSE}
spread_df = na.omit(spread_df)

spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$score_home[spread_df$team_home == spread_df$team_favorite_id] - spread_df$score_away[spread_df$team_home == spread_df$team_favorite_id]

spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$score_home[spread_df$team_away == spread_df$team_favorite_id] - spread_df$score_away[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$score_away[spread_df$team_away == spread_df$team_favorite_id] - spread_df$score_home[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$score_away[spread_df$team_home == spread_df$team_favorite_id] - spread_df$score_home[spread_df$team_home == spread_df$team_favorite_id]

spread_df
```

## Alter diff columns to take spread into account
```{r pressure, echo=FALSE}
spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] + spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]

spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] - spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] + spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] - spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]

spread_df
```


## Create home and away spread value columns
```{r pressure, echo=FALSE}
spread_df$spread_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]
spread_df$spread_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$spread_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]*-1
spread_df$spread_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]*-1

spread_df
```


## Select only important columns so that we can use rbind 
```{r pressure, echo=FALSE}
spread_df = spread_df[, c(1, 7, 9, 4, 8,10)]

spread_df = na.omit(spread_df)
```

## use rbind to condense dataframe into 3 columns 
```{r pressure, echo=FALSE}
home_df = spread_df[, c(1:3)]
away_df = spread_df[, c(4:6)]

away_df_renamed = setNames(away_df, names(home_df))

tidy_df = rbind(home_df, away_df_renamed)

tidy_df = tidy_df %>%
  rename("team" = "team_home",
         "difference" = "diff_home",
         "spread" = "spread_home")

tidy_df
```

## export csv
```{r pressure, echo=FALSE}
write.csv(tidy_df, "tidy_df.csv", row.names = FALSE)
```




## Create new df that contains geographical information for map
```{r pressure, echo=FALSE}
spread_df_whole

map_df = select(spread_df_whole, c(5:8))

#assign a 2 letter state id for every game depending on home team
map_df$location[map_df$team_home == "Atlanta Falcons"] <- "Georgia"
map_df$location[map_df$team_home == "Buffalo Bills"] <- "New York"
map_df$location[map_df$team_home == "Cleveland Browns"] <- "Ohio"
map_df$location[map_df$team_home == "Dallas Cowboys"] <- "Texas"
map_df$location[map_df$team_home == "Green Bay Packers"] <- "Wisconsin"
map_df$location[map_df$team_home == "Kansas City Chiefs"] <- "Missouri"
map_df$location[map_df$team_home == "Miami Dolphins"] <- "Florida"
map_df$location[map_df$team_home == "Minnesota Vikings"] <- "Minnesota"
map_df$location[map_df$team_home == "New England Patriots"] <- "Massachusetts"
map_df$location[map_df$team_home == "New Orleans Saints"] <- "Louisiana"
map_df$location[map_df$team_home == "San Francisco 49ers"] <- "California"
map_df$location[map_df$team_home == "Tennessee Titans"] <- "Tennessee"
map_df$location[map_df$team_home == "Jacksonville Jaguars"] <- "Florida"
map_df$location[map_df$team_home == "Philadelphia Eagles"] <- "Pennsylvania"
map_df$location[map_df$team_home == "New York Jets"] <- "New York"
map_df$location[map_df$team_home == "Indianapolis Colts"] <- "Indiana"
map_df$location[map_df$team_home == "Seattle Seahawks"] <- "Washington"
map_df$location[map_df$team_home == "Pittsburgh Steelers"] <- "Pennsylvania"
map_df$location[map_df$team_home == "Washington Commanders"] <- "Maryland"
map_df$location[map_df$team_home == "Los Angeles Rams"] <- "California"
map_df$location[map_df$team_home == "Arizona Cardinals"] <- "Arizona"
map_df$location[map_df$team_home == "Baltimore Ravens"] <- "Maryland"
map_df$location[map_df$team_home == "Cincinnati Bengals"] <- "Ohio"
map_df$location[map_df$team_home == "Denver Broncos"] <- "Colorado"
map_df$location[map_df$team_home == "Detroit Lions"] <- "Michigan"
map_df$location[map_df$team_home == "Los Angeles Chargers"] <- "California"
map_df$location[map_df$team_home == "Tampa Bay Buccaneers"] <- "Florida"
map_df$location[map_df$team_home == "New York Giants"] <- "New York"
map_df$location[map_df$team_home == "Las Vegas Raiders"] <- "Nevada"
map_df$location[map_df$team_home == "Carolina Panthers"] <- "North Carolina"
map_df$location[map_df$team_home == "Chicago Bears"] <- "Illinois"
map_df$location[map_df$team_home == "Houston Texans"] <- "Texas"

map_df
```


## Create difference column to determine who won the game, ignoring spread
```{r pressure, echo=FALSE}
map_df$diff_home = map_df$score_home - map_df$score_away
map_df$diff_away = map_df$score_away - map_df$score_home

map_df
```

## Use rbind to create 1 dataframe containing team, difference, and location columns
```{r pressure, echo=FALSE}
map_df_home = map_df[, c(1,6,5)]
map_df_away = map_df[, c(4,7,5)]

map_df_away_renamed = setNames(map_df_away, names(map_df_home))

map_df_tidy = rbind(map_df_home, map_df_away_renamed)

map_df_tidy = map_df_tidy %>%
  rename("team" = "team_home",
         "difference" = "diff_home",
         "state" = "location") 
  
map_df_tidy$win[map_df_tidy$difference > 0] = 1
map_df_tidy$win[map_df_tidy$difference <= 0] = 0

map_df_tidy
```

## export csv
```{r pressure, echo=FALSE}
write.csv(map_df_tidy, "map_df_tidy.csv", row.names = FALSE)
```


## create csv for model
```{r pressure, echo=FALSE}
model_df = map_df
nfl_df_2000 = filter(nfl_df, schedule_season > 1999)

model_df$temp1 = nfl_df_2000[, 14]
model_df$temp2 = nfl_df_2000[, 14]

model_df$wind1 = nfl_df_2000[, 15]
model_df$wind2 = nfl_df_2000[, 15]

model_df

model_df_home = model_df[, c(1,6,8,10)]
model_df_away = model_df[, c(4,7,9,11)]

model_df_away_renamed = setNames(model_df_away, names(model_df_home))

model_df_tidy = rbind(model_df_home, model_df_away_renamed)

model_df_tidy = model_df_tidy %>%
  rename("team" = "team_home",
         "difference" = "diff_home",
         "temperature" = "temp1",
         "wind" = "wind1") 

model_df_tidy
```

## export csv
```{r pressure, echo=FALSE}
write.csv(model_df_tidy, "model_df_tidy.csv", row.names = FALSE)
```