---
title: "Data Preprocessing"
output: md_document
---

## Import Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

##Import CSV file
```{r cars}
nfl_df = read.csv('NFL_Spread_Data.csv')

nfl_df
```
## Create New Dataframe Containing only games afte 1999
```{r pressure, echo=FALSE}
spread_df = filter(nfl_df, schedule_season > 1999)

spread_df
```

## Remove unnecessary columns
```{r pressure, echo=FALSE}
spread_df = select(spread_df, c(5:10))
```
## Replace team names that have switched cities or names since 2000
```{r pressure, echo=FALSE}
spread_df$team_home[spread_df$team_home == "Oakland Raiders"] <- "Las Vegas Raiders"
spread_df$team_home[spread_df$team_home == "Washington Redskins"] <- "Washington Commanders"
spread_df$team_home[spread_df$team_home == "Washington Football Team"] <- "Washington Commanders"
spread_df$team_home[spread_df$team_home == "St. Louis Rams"] <- "Los Angeles Rams"
spread_df$team_home[spread_df$team_home == "San Diego Chargers"] <- "Los Angeles Chargers"

spread_df$team_away[spread_df$team_away == "Oakland Raiders"] <- "Las Vegas Raiders"
spread_df$team_away[spread_df$team_away == "Washington Redskins"] <- "Washington Commanders"
spread_df$team_away[spread_df$team_away == "Washington Football Team"] <- "Washington Commanders"
spread_df$team_away[spread_df$team_away == "St. Louis Rams"] <- "Los Angeles Rams"
spread_df$team_away[spread_df$team_away == "San Diego Chargers"] <- "Los Angeles Chargers"
```

## Replace team favorite ids with team names
```{r pressure, echo=FALSE}
team_names = unique(spread_df$team_home)
team_names = append(team_names, "None")

team_ids = c("ATL", "BUF", "CLE", "DAL", "GB", "KC", "MIA", "MIN", "NE", "NO", "NYG", "LV", "PIT", "WAS", "LAR", "ARI", "BAL", "CIN", "DEN", "DET", "IND", "PHI", "LAC", "SF", "SEA", "TB", "TEN", "NYJ", "CAR", "CHI", "JAX", "HOU", "PICK")

names(team_names) = team_ids

spread_df$team_favorite_id = team_names[spread_df$team_favorite_id]

spread_df
```

## Create difference columns that subtracts score home and score away based on the favorite
```{r pressure, echo=FALSE}
spread_df = na.omit(spread_df)

spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$score_home[spread_df$team_home == spread_df$team_favorite_id] - spread_df$score_away[spread_df$team_home == spread_df$team_favorite_id]

spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$score_home[spread_df$team_away == spread_df$team_favorite_id] - spread_df$score_away[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$score_away[spread_df$team_away == spread_df$team_favorite_id] - spread_df$score_home[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$score_away[spread_df$team_home == spread_df$team_favorite_id] - spread_df$score_home[spread_df$team_home == spread_df$team_favorite_id]

spread_df
```

## Alter diff columns to take spread into account
```{r pressure, echo=FALSE}
spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$diff_home[spread_df$team_home == spread_df$team_favorite_id] + spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]

spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$diff_home[spread_df$team_away == spread_df$team_favorite_id] - spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$diff_away[spread_df$team_away == spread_df$team_favorite_id] + spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$diff_away[spread_df$team_home == spread_df$team_favorite_id] - spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]

spread_df
```


## Create home and away spread value columns
```{r pressure, echo=FALSE}
spread_df$spread_home[spread_df$team_home == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]
spread_df$spread_away[spread_df$team_away == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]

spread_df$spread_home[spread_df$team_away == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_away == spread_df$team_favorite_id]*-1
spread_df$spread_away[spread_df$team_home == spread_df$team_favorite_id] = spread_df$spread_favorite[spread_df$team_home == spread_df$team_favorite_id]*-1

spread_df
```


## Select only important columns so that we can use rbind 
```{r pressure, echo=FALSE}
spread_df = spread_df[, c(1, 7, 9, 4, 8,10)]

spread_df
```

## use rbind to condense dataframe into 3 columns 
```{r pressure, echo=FALSE}
home_df = spread_df[, c(1:3)]
away_df = spread_df[, c(4:6)]

away_df_renamed = setNames(away_df, names(home_df))

tidy_df = rbind(home_df, away_df_renamed)

tidy_df = tidy_df %>%
  rename("team" = "team_home",
         "difference" = "diff_home",
         "spread" = "spread_home")

tidy_df
```